Stability and Performance of Coalitions of Prosumers Through Diversification in the Smart Grid

### 2015-11-20 Article.tex
corrections Monique

The recent developments in the smart grid area, such as demand side management mechanisms, storage efficiency, or communication architectures to name a few, sketch a new power system. The rigid top-down centralized architecture of today’s power grids is expected to evolve in more distributed and flexible systems. Bidirectional power flows, storage, and efficient use of information are some of the key principles that will enable this transition. Since production could be located down to the very end of the distribution networks, nodes that were simply pure loads yesterday could behave tomorrow as generators or loads, further complicating energy usages [5]. 
In this paper, we focus on these "nodes" in the distribution network that can produce and consume electricity. More precisely, we consider agents owning both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "prosumer" [6]. A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with better contracts, the operator concerns are related to the quality and reliability of the injections.

@@

We focus on coalitions because it appears as a solution to both the number of prosumers communicating with the operator as well as their high variabilities. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}. On the other hand, it is known that diversification of the assets is a way of minimizing risks when building a portfolio. We thus expect more stable and predictable energy productions for coalitions than for single agents. Nevertheless, all coalitions are not equally stable or productive. Special attention should thus be paid to the aggregation step. Recall that prosumers have both consumption and production components, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}). Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of risk acceptable. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents. Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

Allowing entities using mainly renewables to inject power in the grid is a difficult challenge. On the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Demand side management techniques can indeed mitigate the consumption peaks and waste less production \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing appears as a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the returns distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}. Nevertheless, one of the key point in the Markowitz theory is to consider explicitly the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interesting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

\begin{figure}
\label{fig:process}
	\includegraphics[scale=.5]{Fig2}
	\caption{Process}
\end{figure}

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following (see block 1 of fig. \ref{fig:process}). A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous available extra-production of agent i at time t :

@@

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected productions but presenting a high unpredictability are penalized and can only afford small contracts. The grid operator also specifies a lower bound $ P^{MIN} $ on the contract values as not to overload the market with unrealistic small coalitions. We thus characterized a valid coalition as one satisfying the two conditions :

@@

The Gaussian assumption of this small example is convenient as it allows us to write $ P_{S}^{CRCT \star} $ analytically. Nevertheless, such assumption is rather unrealistic in practice. In the following, we keep the same framework but release this Gaussian assumption unless the contrary is specified (see eq. \ref{eq:alpha_star}). This assumption will indeed be convenient for computing some parameter estimates.

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_3}
\caption{{\footnotesize Histograms of utility values for coalitions of size 3. Red bars stand for cliques in the decorrelation graph, and blue bars for all the other possible triplets. Cliques tend to exhibit higher utilities than random coalitions.}}
\label{fig:histo_cliques}
\end{figure}

Figure \ref{fig:histo_cliques} shows the distributions of the utility values for cliques of size 3 (triangles) in $ G_{2}^{\epsilon^{\star}} $ and for all the other possible triplets of agents. It is clearly visible that cliques tend to exhibit higher utilities because of their de-correlation property. Choosing cliques in $ G_{2}^{\epsilon^{\star}} $ as coalitions seems therefore appealing. Nevertheless, the quality of the results seems to decrease as the sizes of the cliques increase. Indeed, the larger the desired cliques, the more dense $ G_{2}^{\epsilon^{\star}} $ becomes (see equation \ref{epsilon_star}). There is a point where cliques results more from noisy edges than true de-correlation, which decreases the quality of the results. Directly mapping cliques to coalitions by this de-correlation oriented approach is thus not sufficient. It is indeed possible that adding agents to these cliques has the combined effect of increasing the expected production while decreasing its stability. The question revolves around measuring the benefits of this production surplus compared to the disadvantage of having coalition with high volatility. This can be quantified by the marginal benefit in equation \ref{eq:marginal_benefit}.

@@

The first steps consists in computing the de-correlation graph $ G_{2} $ as well as the optimal threshold $ \epsilon^{\star} $. Cliques of size k in $ G_{2}^{\epsilon^{\star}} $ are considered as coalition seeds. The next step is a local greedy improvement over the landscape represented by  $ G_{2}^{\epsilon^{\star}} $. Cliques add alternatively the node $ i^{\star} $ in their neighborhood that yields the best marginal benefit $ MAX_{ i \in N(clique) } \delta_{clique}(i) $ where $ N(clique) $ is the neighborhood of a given clique. See algorithms section in the appendix.

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, we do not have any technical constraints on coalition sizes even if we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents.
\begin{figure}
\includegraphics[scale=.48]{./figs/figure_4}
\caption{{\footnotesize Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to eq. \ref{eq:alpha_star} in order to favor 40 agents coalitions.}}
\label{fig:real_utility2}
\end{figure}

@@

In this paper we studied how aggregations of prosumers could be authorized to sell their surplus of production to the grid operator. By relying on the past values of the agents, we constrained the market entry to both sufficiently productive and stable coalitions. The power that a coalition is able to propose on the market is therefore related to production and stability. The correlations between the prosumers impact directly the volatility of the coalitions, which led us to seek uncorrelated aggregations of agents. We used a graph representation of the correlation relationships between the agents as a reduced landscape for the coalition formation. A greedy algorithm that starts with cliques of the "de-correlation" graph of the agents and makes local improvements offers a good compromise between speed and quality of the results. We compare these results with random samplings, and an opposite strategy that clusters correlated agents together. We showed that the coalitions resulting from our algorithm are able to provide more power to the grid with a lower volatility. Because of their better production over volatility ratios, these coalitions will tend to use less storage and waste less energy than more unstable coalitions. Because in real situations, agents are prone to failure, resilience is also an important criterion for the quality of the aggregations. We therefore studied how the coalitions remain on the market when their agents fail randomly. We showed that, in this situation, the coalitions resulting from our algorithm better withstand losses of agents.

>>>>

The recent developments in the smart grid area, such as demand side management mechanisms, storage efficiency, or communication architectures to name a few, sketch a new power system. The rigid top-down centralized architecture of today’s power grids is expected to evolve in more distributed and flexible systems. Bidirectional power flows, storage, and efficient use of information are some of the key principles that will enable this transition. Since production could be located down to the very end of the distribution networks, nodes that were simply pure loads yesterday could behave tomorrow as generators or loads, further complicating energy usages [5]. 
In this paper, we focus on these "nodes" in the distribution network that can produce and consume electricity. More precisely, we consider agents owning both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "prosumer" [6]. A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the prosumers intend to maximize their benefits with better contracts, the operator concerns are related to the quality and reliability of the injections.

@@

We focus on coalitions because it appears as a solution to both the number of prosumers communicating with the operator as well as their high variabilities. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performances on the communication side \cite{Negeri2012}. On the other hand, it is known that diversification of the assets is a way of minimizing risks when building a portfolio. We thus expect more stable and predictable energy productions for coalitions than for single agents. Nevertheless, all coalitions are not equally stable or productive. Special attention should thus be paid to the aggregation step. Recall that prosumers have both consumption and production components, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of this paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to accept. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}). Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of acceptable risk. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents. Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

Since renewables only produce when the resource is present and 
because these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}, allowing large penetration of renewables in the grid appears as a difficult challenge. One possible solution consists in shifting some of the loads towards high productivity periods. Demand side management techniques can indeed mitigate the consumption peaks and waste less production \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing appears as a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extent.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to accept, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the returns distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}. Nevertheless, one of the key point in the Markowitz theory is to consider explicitly the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interesting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

\begin{figure}
\label{fig:process}
	\includegraphics[scale=.45]{Fig2}
	\caption{Process}
\end{figure}

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following (see block 1 of fig. \ref{fig:process}). A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. We denote by $ P_{i}(t) $ the instantaneous available extra-production of agent i at time t :

@@

This is the best contract value that the coalition S can afford for a given stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected productions but presenting a high unpredictability are penalized and can only afford small contracts. The grid operator also specifies a lower bound $ P^{MIN} $ on the contract values as not to overload the market with unrealistic small coalitions. We thus characterized a valid coalition as one satisfying the two conditions :

@@

The Gaussian assumption of this small example is convenient as it allows us to write $ P_{S}^{CRCT \star} $ analytically. Nevertheless, such an assumption is rather unrealistic in practice. In the following, we keep the same framework but release this Gaussian assumption unless the contrary is specified (see eq. \ref{eq:alpha_star}). This assumption will indeed be convenient for computing some parameter estimates.

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_3}
\caption{{\footnotesize Histograms of utility values for coalitions of size 3. Red bars stand for cliques in the decorrelation graph, and blue bars for all the other possible triples. Cliques tend to exhibit higher utilities than random coalitions.}}
\label{fig:histo_cliques}
\end{figure}

Figure \ref{fig:histo_cliques} shows the distributions of the utility values for cliques of size 3 (triangles) in $ G_{2}^{\epsilon^{\star}} $ and for all the other possible triplets of agents. It is clearly visible that cliques tend to exhibit higher utilities because of their de-correlation property. Choosing cliques in $ G_{2}^{\epsilon^{\star}} $ as coalitions seems therefore appealing. Nevertheless, the quality of the results seems to decrease as the sizes of the cliques increase. Indeed, the larger the desired cliques, the more dense $ G_{2}^{\epsilon^{\star}} $ becomes (see equation \ref{epsilon_star}). There is a point where cliques result more from noisy edges than true de-correlation, which decreases the quality of the results. Directly mapping cliques to coalitions by this de-correlation oriented approach is thus not sufficient. It is indeed possible that adding agents to these cliques has the combined effect of increasing the expected production while decreasing its stability. The question revolves around measuring the benefits of this production surplus compared to the disadvantage of having coalition with high volatility. This can be quantified by the marginal benefit in equation \ref{eq:marginal_benefit}.

@@

The first steps consist in computing the de-correlation graph $ G_{2} $ as well as the optimal threshold $ \epsilon^{\star} $. Cliques of size k in $ G_{2}^{\epsilon^{\star}} $ are considered as coalition seeds. The next step is a local greedy improvement over the landscape represented by  $ G_{2}^{\epsilon^{\star}} $. Cliques add alternatively the node $ i^{\star} $ in their neighborhood that yields the best marginal benefit $ MAX_{ i \in N(clique) } \delta_{clique}(i) $ where $ N(clique) $ is the neighborhood of a given clique. See algorithms section in the appendix.

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, we do not have any technical constraints on coalition sizes though we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents.
\begin{figure}
\includegraphics[scale=.48]{./figs/figure_4}
\caption{{\footnotesize Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to eq. \ref{eq:alpha_star} in order to favor 40 agents coalitions.}}
\label{fig:real_utility2}
\end{figure}

@@

In this paper we studied how aggregations of prosumers could be authorized to sell their surplus of production to the grid operator. By relying on the past values of the agents, we constrained the market entry to both sufficiently productive and stable coalitions. The power that a coalition is able to propose on the market is therefore related to production and stability. The correlations between the prosumers impact directly the volatility of the coalitions, which led us to seek uncorrelated aggregations of agents. We used a graph representation of the correlation relationships between the agents as a reduced landscape for the coalition formation. A greedy algorithm that starts with cliques of the "de-correlation" graph of the agents and makes local improvements offers a good compromise between speed and quality of the results. We compared these results with random samplings, and an opposite strategy that clusters correlated agents together. We showed that the coalitions resulting from our algorithm are able to provide more power to the grid with a lower volatility. Because of their better production over volatility ratios, these coalitions will tend to use less storage and waste less energy than more unstable coalitions. Because in real situations, agents are prone to failure, resilience is also an important criterion for the quality of the aggregations. We therefore studied how the coalitions remain on the market when their agents fail randomly. We showed that, in this situation, the coalitions resulting from our algorithm better withstand losses of agents.
